---
title: "Covid Examples"
output: html_notebook
---

# In-class exercise

```{r}
### Load the libraries will we need
library(readr)
library(dplyr)
library(ggplot2)
```

Read the data and check

```{r}
gapminder <- read_csv("raw_data/gapminder.csv")
head(gapminder)
```
Summarise the gapminder data in an appropriate manner to produce a plot to show the change in average gdpPercap for each continent over time.

```{r}
## advisable to do the summary first and check the result

group_by(gapminder, continent, year) %>% 
  summarise(GDP = mean(gdpPercap))
```
```{r}
group_by(gapminder, continent, year) %>% 
  summarise(GDP = mean(gdpPercap)) %>% 
  ggplot(aes(x = year, y = GDP, fill=continent)) + geom_col() + facet_wrap(~continent)
```


# Some examples of visualising the Covid-19 dataset using `dplyr` and `ggplot2`

First load the required libraries

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

Download the Covid dataset and tidy

```{r}
download.file("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv",destfile = "time_series_covid19_confirmed_global.csv")

covid <- read_csv("raw_data/time_series_covid19_confirmed_global.csv") %>% 
  rename(country = `Country/Region`) %>% 
  pivot_longer(5:last_col(),names_to="Date", values_to="Cases") %>% 
  mutate(Date=as.Date(Date,"%m/%d/%y")) %>% 
  group_by(country,Date) %>% 
  summarise(Cases = sum(Cases))
```

Compare trajectories of different countries. The `%in%` operator is an alternative to or (`|`) to find a country name that can either one of a number of possibilities.

```{r}
filter(covid, country %in% c("United Kingdom","France","Spain")) %>%
  ggplot(aes(x = Date, y = Cases,col=country)) + geom_line()
```


Try to make example similar to [https://www.statista.com/statistics/1110187/coronavirus-incidence-europe-by-country/](https://www.statista.com/statistics/1110187/coronavirus-incidence-europe-by-country/). We're not using up-to-date figures for population so there will be some differences. The two data frames can be joined because they have a column name in common (`country`). Most of the country names that appear in `gapminder` also appear in the `covid` dataset, so not much data is lost in the join. 

Since we are interested in European countries for this dataset, we pre-filter the `gapminder` data. Also, we only need the population values from 2007.


```{r}
##change the date to whichever date you want to look at

filter(gapminder, year == 2007,continent=="Europe") %>% 
  left_join(covid) %>% 
  filter(Date == "2022-01-06") %>% 
  mutate(Cases = round(Cases / (pop / 1e5)))
```

We can use a `geom_col` plot to visualise

```{r}
filter(gapminder, year == 2007,continent=="Europe") %>% 
  left_join(covid) %>% 
  filter(Date == "2022-01-06") %>% 
  mutate(Cases = round(Cases / (pop / 1e5))) %>% 
  ggplot(aes(x = Cases, y = country)) + geom_col()
```

The default ordering for the bars is alphabetical, which is probably not a natural choice. The `forcats` package, which is part of `tidyverse` allows *factors* in a data frame to be re-ordered and re-labeled.

- [forcats package](https://forcats.tidyverse.org/)

In particular, we can use the `fct_reorder` function to reorder the country names according to the number of cases. This can be done inside the `ggplot2` function itself.

```{r}
## this will install the package if it is not already installed
if(!require(forcats)) install.packages(forcats)

filter(gapminder, year == 2007,continent=="Europe") %>% 
  left_join(covid) %>% 
  filter(Date == "2022-01-06") %>% 
  mutate(Cases = round(Cases / (pop / 1e5))) %>% 
  ggplot(aes(x = Cases, y = forcats::fct_reorder(country,Cases))) + geom_col()

```


A heatmap of number of cases over time (similar to that reported by the BBC) can be achieved using a `geom_tile`

```{r}
### Get the 2007 gapminder data to avoid repeating data
filter(gapminder, year ==  2007, continent=="Europe") %>% 
  left_join(covid) %>% 
  filter(!is.na(Cases)) %>% ## remove countries with missing values
  ggplot(aes(x = Date, y = country,fill=Cases)) + geom_tile()
  
```