---
title: "Covid Examples"
output: html_notebook
---

# In-class exercise

```{r}
### Load the libraries will we need
library(readr)
library(dplyr)
library(ggplot2)
```

Read the data and check

```{r}
gapminder <- read_csv("raw_data/gapminder.csv")
head(gapminder)
```
Summarise the gapminder data in an appropriate manner to produce a plot to show the change in average gdpPercap for each continent over time.

```{r}
## advisable to do the summary first and check the result

group_by(gapminder, continent, year) %>% 
  summarise(GDP = mean(gdpPercap))
```
```{r}
group_by(gapminder, continent, year) %>% 
  summarise(GDP = mean(gdpPercap)) %>% 
  ggplot(aes(x = year, y = GDP, fill=continent)) + geom_col() + facet_wrap(~continent)
```


# Some examples of visualising the Covid-19 dataset using `dplyr` and `ggplot2`

First load the required libraries

```{r}
library(readr)
library(dplyr)
library(ggplot2)

```

Download the Covid dataset and tidy

```{r}
download.file("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv",destfile = "time_series_covid19_confirmed_global.csv")

covid <- read_csv("time_series_covid19_confirmed_global.csv") %>% 
  rename(country = `Country/Region`) %>% 
  tidyr::gather(Date, Cases, 5:last_col()) %>% 
  mutate(Date=as.Date(Date,"%m/%d/%y"))
```

Finding the European countries with most cases

```{r}
gapminder <- read_csv("raw_data/gapminder.csv")
filter(gapminder, year ==  2007, continent=="Europe") %>% 
  left_join(covid) %>% 
  filter(Date == "2020-06-04") %>% 
  group_by(country) %>% 
  summarise(Cases = sum(Cases)) %>% 
  arrange(desc(Cases))

```

Compare trajectories of different countries

```{r}
filter(covid, country %in% c("United Kingdom","France","Spain")) %>% 
  group_by(country,Date) %>% 
  summarise(Cases = sum(Cases)) %>% 
  ggplot(aes(x = Date, y = Cases,col=country)) + geom_line()
```

Heatmap of number of cases over time

```{r}
### Get the 2007 gapminder data to avoid repeating data
filter(gapminder, year ==  2007, continent=="Europe") %>% 
  left_join(covid) %>% 
  group_by(country,Date) %>% 
  summarise(Cases = sum(Cases)) %>% ## sum in case a country has multiple rows (e.g. United Kingdom)
  filter(!is.na(Cases)) %>% ## remove countries with missing values
  ggplot(aes(x = Date, y = country,fill=Cases)) + geom_tile()
  
```
Try to make example similar to [https://www.statista.com/statistics/1110187/coronavirus-incidence-europe-by-country/](https://www.statista.com/statistics/1110187/coronavirus-incidence-europe-by-country/). We're not using up-to-date figures for population so there will be some differences. Also not all countries appear in gapminder

```{r}
filter(gapminder, year == 2007,continent=="Europe") %>% 
  left_join(covid) %>% 
  filter(Date == "2021-02-14") %>% 
  mutate(Cases = round(Cases / (pop / 1e5))) %>% 
  arrange(desc(Cases)) %>% 
  select(country, Cases)
```


Compare average number of cases per-continent

```{r}
filter(gapminder, year ==  2007) %>% 
  left_join(covid) %>% 
  group_by(continent,Date) %>% 
  summarise(Cases = mean(Cases)) %>% 
  ggplot(aes(x = Date, y = Cases, fill=continent))+ geom_col() + facet_wrap(~continent,scale="free_y")
```


